Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-08 paper/main.tex
Update

Figure~\ref{fig:intro} highlights some of these issues. To assist researchers
in constructing DAGs manually, we propose an iterative model construction
method that combines domain knowledge with data-driven insights. Our approach
utilizes conditional independence testing to identify pair of variables in the
DAG that are missing an edge or have a superfluous edge. Based on this
information, in each iteration we present one such pair of variables to an
expert who then decides the orientation of the edge between the variables based
on their domain knowledge. The method also iteratively prunes all superfluous
edges in each iteration. As a result this approach guides the expert in
constructing their model while ensuring that the model is consistent with the
data.

@@

\begin{enumerate}
    \item We propose a novel measure of conditional association for mixed data
	    based on canonical correlations
	    (Section~\ref{sec:mixed_association}). This measure generalizes
	    several commonly used special case measures of association to mixed
	    data.
    \item Using this measure of conditional association, we develop a procedure
	    to rank violations of implied CIs. Using this ranking helps us
	    prioritize modifications to DAG to improve its fit while allowing
	    us to integrate expert knowledge when adding new edges (Section~\ref{sec:modification}).
    \item We provide a web-based interactive tool and a Python package to allow 
	    researchers to easily apply this method to their datasets (Section~\ref{sec:web}).
    \item Lastly, we compare the manual DAG construction method with automated
	    causal discovery algorithms (Section~\ref{sec:empirical}).
\end{enumerate}

@@

\subsection{Ranking Potential New Edges}
\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{RankedExpand}{RankedExpand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc RankedExpand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$\phi$,$B=\emptyset$)}{
    let $\phi_{\max} \gets 0 $\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
		\If{$\lvert \phi(X, Y, \mathbf{Z}) \rvert > \phi_{\max}$}{
			$ \phi_{\max} \gets \lvert \phi(X, Y, \mathbf{Z}) \rvert $ \;
			$ L_{\max} \gets \mathcal{A}_G(X, Y)$ \;
		}
        }
    }
    add $L_{\max}$ to $E$\;
    \Return{$(V,E)$}\;
}
\caption{Adding an edge between variables with the highest correlation}
\label{algo:ranked_expand}
\end{algorithm}

>>>>

Figure~\ref{fig:intro} highlights some of these issues.   When manually
constructing models, it is important to test whether the resulting DAG
accurately represents the dataset. Based on this evaluation and domain
knowledge, we can make further modification to the model. One common approach
for assessing DAGs is to check whether the Conditional Independences (CIs)
implied by the DAG hold in the data \citep{Ankan2021}. If we find violations to
these CIs, we can use our domain knowledge to make appropriate changes to the
model.

In this paper we present a method that utilizes this idea of CI testing to
assist researchers during the DAG construction instead of testing the final
model. We propose an iterative model construction method that combines domain
knowledge with data-driven insights. Our approach utilizes conditional
independence testing to identify pair of variables in the DAG that are missing
an edge or have a superfluous edge. Based on this information, in each
iteration we present one such pair of variables to an expert who then decides
the orientation of the edge between the variables based on their domain
knowledge. The method also iteratively prunes all superfluous edges in each
iteration. As a result this approach guides the expert in constructing their
model while ensuring that the model is consistent with the data.

@@

\begin{enumerate}
%     \item We propose a novel measure of conditional association for mixed data
% 	    based on canonical correlations
% 	    (Section~\ref{sec:mixed_association}). This measure generalizes
% 	    several commonly used special case measures of association to mixed
% 	    data.
    \item We develop an iterative method based on CI testing to assist researchers in 
	    constructing DAGs using their domain knowledge (Section~\ref{sec:modification}).
    \item We develop a ranking method to prioritize potential modifications in the DAG that 
    	    can help in improving the model's fit quickly (Section~\ref{sec:ranking}).
    \item We provide a web-based interactive tool and a Python package to allow 
	    researchers to easily apply this method to their datasets (Section~\ref{sec:web}).
    \item Lastly, we compare the manual DAG construction method with automated
	    causal discovery algorithms (Section~\ref{sec:empirical}).
\end{enumerate}

@@

\subsection{Ranking Potential New Edges}
\label{sec:ranking}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{RankedExpand}{RankedExpand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc RankedExpand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$\phi$,$B=\emptyset$)}{
    let $\phi_{\max} \gets 0 $\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
		\If{$\lvert \phi(X, Y, \mathbf{Z}) \rvert > \phi_{\max}$}{
			$ \phi_{\max} \gets \lvert \phi(X, Y, \mathbf{Z}) \rvert $ \;
			$ L_{\max} \gets \mathcal{A}_G(X, Y)$ \;
		}
        }
    }
    add $L_{\max}$ to $E$\;
    \Return{$(V,E)$}\;
}
\caption{Adding an edge between variables with the highest correlation}
\label{algo:ranked_expand}
\end{algorithm}
